---
title: "Data cleaning pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data cleaning pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

This vignette demonstrates how the data-cleaning functions in adRutils work together on a realistically messy dataset. The functions are designed to work together seamlessly: whenever possible, they add new columns rather than mutating existing ones, making transformations easy to inspect and reverse. The example builds a pipeline that addresses missing data, duplicates, outliers, longitudinal coalescing, and binning—roughly in the order these issues are typically encountered.

```{r setup}
library(adRutils)
library(dplyr)
```

## A messy dataset

We'll simulate a study dataset with the kinds of problems that actually show up: duplicate rows from a faulty join, columns that are nearly all `NA`, outliers from data-entry errors, the same measure collected at multiple visits, and a continuous variable that needs binning for downstream analysis.

```{r simulate}
set.seed(7)

n <- 100
raw_data <- tibble::tibble(
  subject_id     = c(sprintf("S%03d", 1:n), "S042", "S073"),  # two duplicates
  age            = c(round(rnorm(n, 65, 8), 1), 67, 71),
  bmi            = c(round(rnorm(n, 26, 4), 1), 27, 24),
  weight_kg      = c(round(rnorm(n, 75, 12), 1), 250, 78),    # one outlier
  almost_empty   = c(rep(NA_real_, n - 1), 1.5, NA, NA),
  fully_empty    = rep(NA_real_, n + 2),
  dsst_v1        = round(rnorm(n + 2, 50, 10)),
  dsst_v2        = ifelse(runif(n + 2) > 0.4, round(rnorm(n + 2, 52, 10)), NA),
  visit_year     = sample(c(2020, 2021), n + 2, replace = TRUE)
)

# inject a few NAs in dsst_v1 too
raw_data$dsst_v1[sample(seq_len(nrow(raw_data)), 15)] <- NA

dim(raw_data)
```

## Step 1: inspect missingness

Always look at NA patterns before you delete anything. `summarize_na()` gives you a sorted view; `drop_high_na_cols()` removes columns above a threshold.

```{r summarize-na}
summarize_na(raw_data)
```

`fully_empty` is 100% missing and `almost_empty` is close to it. We can drop both with a threshold of, say, 90%:

```{r drop-na-cols}
data <- drop_high_na_cols(raw_data, threshold = 90)
names(data)
```

`drop_high_na_cols()` prints what it removed; if you want the same view without modifying the data, call `summarize_na(raw_data, threshold = 90)`.

## Step 2: deduplicate

Our synthetic data has two duplicated subject IDs (`S042`, `S073`). `remove_duplicates()` defaults to keeping the row with the fewest NAs per duplicate ID — usually what you want when one record is more complete than the other.

```{r dedup}
data <- remove_duplicates(data, subject_id)
nrow(data)
```

The other strategies are `keep = "first"`, `"last"`, and `"none"` (which removes *all* duplicates including the originals — useful when you suspect any duplicated ID is a data integrity problem).

## Step 3: combine longitudinal visits

`dsst_v1` and `dsst_v2` are the same measure collected at two visits. Often you want a single coalesced column with a preferred source, plus a tracking column showing which visit contributed for each row.

```{r combine-tp}
data <- combine_timepoints(
  data,
  base_names = "dsst",
  timepoints = c("_v1", "_v2"),
  timepoint_labels = c("_v1" = "Visit 1", "_v2" = "Visit 2")
)

data %>%
  select(subject_id, dsst_v1, dsst_v2, dsst_final, dsst_source) %>%
  head()
```

`dsst_final` takes the first non-missing value across visits in the priority order you specified; `dsst_source` records which visit it came from. The original `dsst_v1` and `dsst_v2` are preserved unchanged.

## Step 4: handle outliers

Our `weight_kg` column has a 250 kg value that's almost certainly a data-entry error. `replace_outliers_with_na()` uses the IQR rule (Tukey's 1.5×IQR by default) to identify and replace outliers with `NA`.

```{r outliers}
summary(data$weight_kg)

data <- replace_outliers_with_na(
  data,
  var_names  = "weight_kg",
  multiplier = 1.5
)

summary(data$weight_kg)
```

Two things worth knowing about how this works:

1. The function is **idempotent** — running it twice produces the same result, since outliers become `NA` and `NA` is skipped on subsequent passes.
2. If you have *paired* columns (e.g., `weight` and `log_weight`), the `paired_cols` argument nullifies the paired entry whenever the source column has an outlier, keeping derived columns consistent with their source.

If you only want to inspect the cutoffs without replacing anything, `detect_outlier_thresholds()` returns the lower and upper bounds:

```{r thresholds}
detect_outlier_thresholds(data, "weight_kg", multiplier = 1.5)
```

## Step 5: bin continuous variables

Many downstream analyses want categorical age groups, BMI categories, etc. `bin_and_categorize_variables()` handles three transformation types — cutpoints (continuous), categorical mappings, and custom functions — through a single interface.

```{r bin}
data <- bin_and_categorize_variables(data, groups = list(
  list(
    col       = "age",
    type      = "cutpoints",
    cutpoints = c(60, 75),
    labels    = c("Younger", "Middle", "Older"),
    name      = "age_group"
  ),
  list(
    col       = "bmi",
    type      = "cutpoints",
    cutpoints = c(18.5, 25, 30),
    labels    = c("Underweight", "Normal", "Overweight", "Obese"),
    name      = "bmi_category"
  )
))

data %>%
  count(age_group, bmi_category)
```

A note on bin semantics: cutpoints are interpreted with `cut()`'s left-open right-closed rule, except the leftmost interval is closed on both ends (matching `include.lowest = TRUE`). So with cutpoints `c(18.5, 25, 30)`, a BMI of exactly 25 falls into `Normal`, not `Overweight`.

## Putting it together

The whole pipeline reads cleanly as a single chain:

```{r pipeline, eval = FALSE}
clean_data <- raw_data %>%
  drop_high_na_cols(threshold = 90) %>%
  remove_duplicates(subject_id) %>%
  combine_timepoints(
    base_names = "dsst",
    timepoints = c("_v1", "_v2")
  ) %>%
  replace_outliers_with_na(var_names = "weight_kg") %>%
  bin_and_categorize_variables(groups = list(
    list(col = "age", type = "cutpoints", cutpoints = c(60, 75)),
    list(col = "bmi", type = "cutpoints", cutpoints = c(18.5, 25, 30))
  ))
```

## What to do next

After cleaning, the typical next steps are:

- Inspect the cleaned dataset with `summarize_na()` again to confirm nothing unexpected is missing.
- Convert pattern-named columns to factors with `convert_columns_to_factors()` if you have many of them.
- Pull related-but-named-differently columns together with `coalesce_variables()` if your dataset uses inconsistent naming across sources.
- Move on to modeling with `fit_models_by_group()` — see the meta-analysis vignette for a worked example.
