---
title: "Meta-analysis workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meta-analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

This vignette walks through a full meta-analysis pipeline in `adRutils`: aligning two cohorts, fitting parallel models in each, extracting tidy coefficients, and pooling effects via REML. The functions involved are designed to work together seamlessly, but you might not see that from any single reference page — the goal here is to show the pipeline end to end on a small synthetic example.

```{r setup}
library(adRutils)
library(dplyr)
```

## A two-cohort design

We'll simulate a discovery cohort and a replication cohort, each with the same outcomes and predictors, but with slightly different sample sizes and effect magnitudes. This mirrors a common pattern in genetic and biomarker studies: estimate effects in one population, then test whether they hold in another.

```{r simulate-data}
set.seed(42)

simulate_cohort <- function(n, cohort_label, effect_age, effect_treatment) {
  tibble::tibble(
    subject_id = paste0(cohort_label, "_", seq_len(n)),
    cohort     = cohort_label,
    age        = round(rnorm(n, mean = 65, sd = 8), 1),
    sex        = sample(c("F", "M"), n, replace = TRUE),
    treatment  = factor(sample(c("control", "active"), n, replace = TRUE),
                        levels = c("control", "active")),
    nfl = effect_age * scale(age)[, 1] +
                  effect_treatment * (treatment == "active") +
                  rnorm(n, sd = 1),
    ab42 = 0.5 * scale(age)[, 1] +
                  0.2 * (treatment == "active") +
                  rnorm(n, sd = 1)
  )
}

discovery   <- simulate_cohort(n = 250, "discovery",   effect_age = 0.40, effect_treatment = 0.55)

replication <- simulate_cohort(n = 180, "replication", effect_age = 0.35, effect_treatment = 0.45)

combined <- bind_rows(discovery, replication)

head(combined)
```

## Step 1: pre-flight alignment check

Before pooling, it's worth confirming that the two cohorts share the same set of group/term combinations you plan to analyze. `check_cohort_alignment()` flags any combinations present in only one cohort.

```{r alignment-check}
check_cohort_alignment(
  discovery, replication,
  by      = c("treatment", "sex"),
  x_label = "discovery",
  y_label = "replication"
)
```

A perfect match is the happy path. In a real analysis, this step often catches issues like a category coded `"Control"` in one cohort and `"control"` in the other, or a level present in only one site.

## Step 2: fit models in each cohort

`fit_models_by_group()` fits a linear model for every (outcome, group) combination. By passing `group_col = "cohort"`, we get one model per outcome per cohort — exactly what we need before pooling.

```{r fit-models}
results <- fit_models_by_group(
  data            = combined,
  outcomes        = c("nfl", "ab42"),
  base_predictors = c("age", "sex", "treatment"),
  group_col       = "cohort",
  groups          = c("discovery", "replication"),
  quiet           = TRUE
)

results %>%
  select(outcome, cohort, model_equation, n_obs)
```

The returned tibble has list-columns for the fitted model object (`model_obj`) and tidy results (`model_res`), plus the data subset used (`dataf`). For pooling, we only need the tidy coefficients.

## Step 3: extract tidy coefficients

`extract_coefficients()` unnests `model_res` and adds confidence intervals computed from the standard errors. We can filter to a specific term — here, the treatment effect.

```{r extract-coefs}
treatment_effects <- extract_coefficients(
  results,
  term_pattern = "treatmentactive"
)

treatment_effects %>%
  select(outcome, cohort, term, estimate, std.error, conf.low, conf.high, p.value)
```

We now have one row per cohort per outcome — the input shape `add_meta_pooled_results()` expects.

## Step 4: REML pooling

`add_meta_pooled_results()` runs random-effects meta-analysis (REML via `metafor::rma()`) within each group defined by `group_vars`, appends a pooled row, and adds a `weight` column showing each cohort's percent contribution to the pooled estimate.

```{r pool}
pooled <- add_meta_pooled_results(
  treatment_effects,
  group_vars = c("outcome", "term")
)

pooled %>%
  select(outcome, cohort, estimate, std.error, conf.low, conf.high,
         weight, I2, same_direction, replicated)
```

A few things to notice in the output:

- **`weight`** is filled in for cohort rows (REML contribution as a percent) and `NA` on the pooled row.
- **`I2`, `Q`, `tau2`** are heterogeneity statistics, attached only to the pooled row.
- **`same_direction`** is `TRUE` when the discovery and replication estimates have the same sign.
- **`replicated`** is `TRUE` when (a) the signs agree *and* (b) the replication estimate has |β/SE| > 1.96.

## Interpreting `replicated`

The `replicated` flag is a useful heuristic but not a formal test. It captures the common-sense idea that a finding replicates if it points the same direction in both cohorts and reaches roughly nominal significance in the second. Two caveats:

1. It only applies when k = 2. For k ≥ 3 cohorts, sign agreement and "the replication estimate" stop being well-defined concepts, so both columns are set to `NA`. If you have three or more cohorts, lean on the pooled p-value, `I2`, and the cohort-level CIs instead.
2. The 1.96 threshold treats the replication cohort as a *separate* significance test rather than building a joint replication statistic. Methods like the Sidak-corrected replication test are stricter; this flag is best read as "directionally consistent and individually significant in the replication."

## What about k ≥ 3?

If you have more than two cohorts, the same pipeline works — `add_meta_pooled_results()` will still run REML and report `weight`, `I2`, `Q`, and `tau2` — but `same_direction` and `replicated` will be `NA` on the pooled rows. Heterogeneity (`I2`) becomes the more informative signal at that point: low `I2` with a tight pooled CI is the multi-cohort analog of a clean replication.

## Recap

The pipeline is:

```
check_cohort_alignment()        # pre-flight
  ↓
fit_models_by_group()           # parallel fits
  ↓
extract_coefficients()          # tidy + CIs
  ↓
add_meta_pooled_results()       # REML pool
```

Each step takes the previous step's output as-is. If you want to filter or transform between steps (e.g., drop unstable estimates before pooling), do it on the tidy frame after `extract_coefficients()` — that's the most ergonomic point in the pipeline to intervene.
