---
title: "Trajectory classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectory classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

`classify_trajectory_groups()` extracts random slopes from a fitted `lme4` model, z-scores them, and assigns each subject to a trajectory group: low / mid / high. 

This vignette walks through the typical use case — taking longitudinal data, fitting a mixed-effects model with random slopes, and using the resulting groups as a downstream predictor — and compares the four available classification methods on the same data.

```{r setup}
library(adRutils)
library(dplyr)
library(lme4)
```

## Why classify trajectories?

Many longitudinal studies want to ask: *who is declining faster than expected?* The natural quantity is the random slope from a mixed-effects model — each subject's deviation from the overall trend. But a continuous slope estimate isn't directly usable as a group label, and it isn't directly comparable across studies. Discretizing into low / mid / high gives you a categorical predictor you can drop into downstream models, plot stratified by group, and report in a table.

This is a conceptual simplification and you should think about whether it's appropriate for your analysis — the slope is a noisy estimate, especially for subjects with few visits, and binning loses information. But when the goal is "find the decliners and describe what they look like," it's a reasonable first move.

## Simulate longitudinal data

We'll simulate 200 subjects with 3–6 visits each, a shared linear trend, and subject-specific random intercepts and slopes.

```{r simulate}
set.seed(2024)

n_subjects <- 200

subjects <- tibble::tibble(
  subject_id      = sprintf("sub_%03d", seq_len(n_subjects)),
  baseline_age    = round(rnorm(n_subjects, 70, 6), 1),
  random_intercept = rnorm(n_subjects, 0, 1),
  random_slope    = rnorm(n_subjects, -0.4, 0.3)   # average decline of -0.4 / yr
)

long_df <- subjects %>%
  rowwise() %>%
  mutate(
    n_visits = sample(3:6, 1),
    visits   = list(sort(round(c(0, runif(n_visits - 1, 1, 5)), 2)))
  ) %>%
  tidyr::unnest(visits) %>%
  rename(time = visits) %>%
  mutate(
    score = 50 +
      random_intercept +
      (-0.4 + random_slope) * time +
      rnorm(n(), 0, 1)
  ) %>%
  select(subject_id, baseline_age, time, score) %>%
  ungroup()

head(long_df, 10)
```

## Fit a mixed-effects model

Standard random-intercepts-and-slopes specification: each subject gets their own intercept and their own slope on `time`.

```{r fit}
fit <- lmer(score ~ time + (1 + time | subject_id), data = long_df)
summary(fit)$coefficients
```

The fixed effect on `time` recovers the average decline (~-0.4). What we want for classification is the per-subject *deviation* from that average, which lives in the random effects.

## Classify with all four methods

`classify_trajectory_groups()` accepts any subset of `"1sd"`, `"1.5sd"`, `"tertile"`, and `"quartile"`. Asking for all four lets us compare them side by side.

```{r classify}
trajectories <- classify_trajectory_groups(
  fit,
  id_col     = "subject_id",
  slope_term = "time",
  methods    = c("1sd", "1.5sd", "tertile", "quartile"),
  labels     = list(low = "Decliner", mid = "Stable", high = "Improver"),
  prefix     = "traj"
)

head(trajectories)
```

The output has one row per subject, with the random intercept (when the model has one), the raw random slope, the z-scored slope, and one factor column per requested method.

## How the methods differ

The four methods answer subtly different questions:

- **SD-based (`1sd`, `1.5sd`)** thresholds the *z-scored slope*. Subjects more than 1 (or 1.5) SDs from the group mean are flagged. The cutoffs are anchored to the spread of the data — fewer subjects fall in the extreme groups when the distribution is tight.
- **Quantile-based (`tertile`, `quartile`)** thresholds the *raw slope* and guarantees a fixed share in each bucket: ~33% / 34% / 33% for tertiles, ~25% / 50% / 25% for quartiles.

A quick comparison of group sizes:

```{r compare-methods}
trajectories %>%
  count(traj_1sd) %>%
  rename(group = traj_1sd, n_1sd = n) %>%
  full_join(
    trajectories %>% count(traj_1.5sd) %>% rename(group = traj_1.5sd, n_1.5sd = n),
    by = "group"
  ) %>%
  full_join(
    trajectories %>% count(traj_tertile) %>% rename(group = traj_tertile, n_tertile = n),
    by = "group"
  ) %>%
  full_join(
    trajectories %>% count(traj_quartile) %>% rename(group = traj_quartile, n_quartile = n),
    by = "group"
  )
```

`1sd` flags ~32% of subjects as extreme (16% each tail); `1.5sd` is much stricter. Tertiles guarantee roughly equal-sized groups regardless of the slope distribution.

**Which to pick?** It depends on the downstream question:

- If you want roughly balanced groups for stratified plotting or cross-tabulation: tertiles or quartiles.
- If you want to flag *unusual* trajectories where "unusual" is defined relative to the cohort's spread: `1sd` or `1.5sd`.
- If you're not sure, run two methods and check that your downstream conclusions don't hinge on the choice. Stable conclusions are more trustworthy than ones that flip between tertile and 1.5sd.

## Using groups as downstream predictors

The whole point of classifying is usually to feed the groups into something else. Here we'll merge the trajectory labels back to a subject-level table and use `fit_models_by_group()` to look at how baseline age varies by trajectory.

```{r downstream}
subject_level <- long_df %>%
  group_by(subject_id, baseline_age) %>%
  summarise(.groups = "drop") %>%
  left_join(trajectories, by = "subject_id")

age_by_traj <- fit_models_by_group(
  data            = subject_level,
  outcomes        = "baseline_age",
  base_predictors = "traj_tertile",
  quiet           = TRUE
)

extract_coefficients(age_by_traj)
```

The `traj_tertile` coefficients are contrasts against the reference level (`Stable` by default — `classify_trajectory_groups()` puts `mid` first because most analyses contrast extremes against typical).

## A warning sign

If `lme4::isSingular(fit)` is `TRUE`, or if the model summary reports a near-zero variance on the slope term, the random slopes are unstable and so are any classifications based on them. `classify_trajectory_groups()` will warn in this case but still return a result — the warning is your cue to revisit the model, not to silently use the output.

```{r singular-check}
isSingular(fit)
```

If you do see singularity, common fixes include simplifying the random-effects structure, increasing the number of visits per subject (if you have control over data collection), or accepting that trajectory classification isn't well-supported by your data.

## Recap

The full pipeline is:

```
fit lmer with (1 + time | subject)
  ↓
classify_trajectory_groups()
  ↓
join groups back to subject-level data
  ↓
fit_models_by_group() / plotting / tables
```

Most of the complexity is in choosing the classification method and reading the random-effects diagnostics — `classify_trajectory_groups()` itself just bundles the conventional steps (extract, z-score, threshold) into one call.
