---
title: "Log-Transformation and Z-Scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-Transformation and Z-Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

Many analyses need numeric measures put on a common, comparable scale before
modeling. The usual sequence is **handle outliers → transform skewed variables →
standardize**, and adRutils provides a function for each step, all of which add
new columns rather than overwriting originals so every stage is easy to inspect
and reverse.

This pattern applies to any set of skewed, differently-scaled numeric measures —
assay panels, lab chemistries, gene-expression values, questionnaire subscales,
and so on. We use a small set of plasma biomarkers as the running example here
because they exercise every step (right-skewed distributions on very different
numeric scales), but nothing in the workflow is specific to biomarkers. The result is 
a set of standardized variables (mean 0, SD 1) suitable for comparison or combining;
the outlier and log steps are preparation that makes that final standardization more robust.

The [Data cleaning pipeline](data_cleaning.html) vignette covers the upstream
steps — missingness, duplicates, binning. This one picks up at the
transform-and-standardize stage.

```{r setup, message = FALSE}
library(adRutils)
library(dplyr)
```

## An example dataset

Three measures on very different scales: `marker_a` in the hundreds, `marker_b`
and `marker_c` both small and right-skewed.

```{r data}
set.seed(7)
n <- 60

measures <- tibble::tibble(
  subject_id = sprintf("S%03d", 1:n),
  ancestry   = sample(c("EUR", "AFR"), n, replace = TRUE),
  marker_a   = round(rnorm(n, 900, 150), 1),
  marker_b   = round(rlnorm(n, log(2),  0.5), 2),
  marker_c   = round(rlnorm(n, log(15), 0.6), 1)
)

# inject one implausible value to demonstrate outlier handling
measures$marker_c[1] <- 500

head(measures)
```

## Step 1: Handle outliers

`replace_outliers_with_na()` flags values outside the Tukey IQR fence and sets
them to `NA`. Apply it to all three measures at once.

```{r outliers}
measures <- replace_outliers_with_na(
  measures,
  var_names          = c("marker_a", "marker_b", "marker_c"),
  multiplier         = 1.5,
  remove_all_na_rows = FALSE
)
```

To inspect the cutoffs without altering the data, `detect_outlier_thresholds()`
returns the bounds for a single column:

```{r thresholds}
detect_outlier_thresholds(measures, "marker_c", multiplier = 1.5)
```

## Step 2: Transform skewed variables

`transform_log10()` adds `log10_<var>` columns and leaves the originals in place.
It requires strictly positive input, which suits concentration-style measures.

```{r log}
measures <- transform_log10(
  measures,
  vars = c("marker_a", "marker_b", "marker_c")
)

names(measures)
```

## Step 3: Standardize

Standardizing the log-scaled measures to z-scores (mean 0, SD 1) puts them on a
common scale so they can be compared or combined. We standardize the `log10_`
columns from Step 2.

```{r zscores}
log_vars <- c("log10_marker_a", "log10_marker_b", "log10_marker_c")

measures <- compute_zscores(
  measures,
  vars   = log_vars,
  prefix = "z_"
)

measures %>%
  select(subject_id, dplyr::starts_with("z_")) %>%
  head()
```

When distributions differ systematically across a grouping variable,
standardizing within strata keeps each group internally comparable. Pass
`group_vars`:

```{r zscores-grouped}
measures <- compute_zscores(
  measures,
  vars       = log_vars,
  prefix     = "z_within_ancestry_",
  group_vars = "ancestry"
)
```

## The pipeline together

```{r pipeline, eval = FALSE}
standardized <- raw_measures %>%
  replace_outliers_with_na(
    var_names          = c("marker_a", "marker_b", "marker_c"),
    remove_all_na_rows = FALSE
  ) %>%
  transform_log10(vars = c("marker_a", "marker_b", "marker_c")) %>%
  compute_zscores(
    vars   = c("log10_marker_a", "log10_marker_b", "log10_marker_c"),
    prefix = "z_"
  )
```

## What to do next

The standardized `z_` columns are ready for modeling — for example fitting models
across groups with `fit_models_by_group()` ([REML Meta-analysis](reml_meta_analysis.html)), or
building composite indices from the standardized values.

## Session information

```{r session}
sessionInfo()
```
